On the Programmatic Generation of Reproducible Documents

### 2020-06-23 inst/jss/kane-listdown/listdown-jss.tex
Renaming.

---
documentclass: jss
author:
  - name: Michael Kane
    affiliation: Yale University
    address:
    - 60 College Street
    - New Haven, CT 06510, USA
    email: michael.kane@yale.edu
  - name: Simon Urbanek
    affiliation: The University of Aukland
    address:
    - 38 Princes Street
    - Auckland Central, Aukland 1010, NZ
    email: s.urbanek@auckland.ac.nz
title:
  formatted: "On the Programmatic Generation of Reproducible Documents"
  # If you use tex in the formatted title, also supply version without
  plain:     "On the Programmatic Generation of Reproducible Documents"
  # For running headers, if needed
  short:     "On the Programmatic Generation of Reproducible Documents"
abstract: >
  Reproducible document standards, like \proglang{R} Markdown, facilitate the programmatic creation of documents whose content is itself programmatically generated. While these documents are generally not complete in the sense that they will not include prose content, generated by an author to provide context, a narrative, etc., programmatic generation can provide substantial efficiencies for structuring and constructing documents. This paper explores the programmatic generation of reproducible by distinguishing components than can be created by computational means from those requiring human-generated prose, providing guidelines for the generation of these documents, and identifying a use case in clinical trial reporting. These concepts and use case are illustrated through the \pkg{listdown} package for the \proglang{R} programming environment, which is is currently available on the Comprehensive R Archive Network (CRAN).
keywords:
  # at least one keyword must be supplied
  formatted: [reproducibility, reproducible documents, data presentation]
  plain:     [reproducibility, reproducible documents, data presentation]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

# Background and concepts

\proglang{R} Markdown [@baumer2014] facilitates the construction of computationally reproducible documents by allowing authors to insert \proglang{R} code for data processing, exploration, analysis, table-making, and visualization directly into structured, electronic documents. The resulting documents are made up of these chunks of \proglang{R} code, which we will refer to as _computational components_ since they are generated by computational means, as well as _narrative components_, which (in scientific writing) is prose intended to contextualize computational components, provide background, define goals, establish themes, and convey the results of the document.
The integration of narrative and computational components was identified as "Literate Programming" by @knuth1984 and software tools, like Sweave [@leisch2002], have supported this functionality for almost two decades.

\proglang{R} Markdown is particularly popular with its success likely being driven by two factors. The first is the relative ease with which these documents can be constructed. While \LaTeX \ is more expressive, it is relatively technical and requires an investment in time to become proficient. By contrast \proglang{R} Markdown documents are easier to create and format and, when the document is used to create \LaTeX, formatting can be passed through to the underlying .tex file. The second factor driving adoption is likely its support for creating _modifiable_ documents, namely Microsoft Word documents. Researchers and analysts, especially those creating applied statistical analyses, often collaborate with domain experts with less technical knowledge. In these cases, the analyst focuses on creating the computational components and narrative components related to results and interpretation. After this initial document is created, the domain expert is free to develop narrative components directly in the document without needing to go through the analyst.

Since computational components are, by definition, computationally derived objects and \proglang{R} Markdown is a well-defined standard, it is possible to programmatically create \proglang{R} Markdown documents with computational components, which is the focus of this paper. Generating documents in this manner has two appealing characteristics. First, it allows us to distinguish the presentation of analytical results from other steps in a data science or data processing pipeline. These other steps including cleaning and analysis often require their own environment and configuration with requirements very different than the computational needs of creating a presentation. By separating these components each can be developed independently. At the same time, by specifying a contract for the output of those objects, we can establish a consistent means by which processed data can be passed to systems for presentating those data in a structured way. The second reason for programmatic creation of R Markdown documents is convenience. In collaborative environments, especially in the early stages, large numbers of graphs and tables are generated and discussed. By collecting these artifacts and structuring them consistently, we can quickly iterate upon and restructure the resulting documents to more clearly present the data without needing to spend time on the creation of the presentation document.

The \pkg{listdown} package provides functions to programmatically create \proglang{R} Markdown files from named lists. It is intended for data analysis pipelines where the presentation of the results
is separated from their creation. For this use case, a data processing (or analysis) is performed
and the results are provided in a single named list, organized heirarchically. List element names denote sections, subsections, subsubsection, etc. and the list elements contain the data structure to be presented including graphs and tables.  The package has native support for \pkg{workflowr} [@blischak2019], pdf, word, or html document along with functions allowing a user to easily extend to other types of supported documents. The goal of the package is to create a documents with all tables and visualization that will appear (computational components). This serves as a starting point from which a user can organize outputs, describe a study, discuss results, and provide conclusions (narrative components).

@@

In this case, data presentation can be fixed across trials. This is especially true in the initial stages, which focus on patient demographics and enrollment. This approach has made it convenient for our group to quickly generate standardize and complete reports for multiple trials concurrently. To date, we have used listdown to report on five clinical trials, with another two currently in process. Results are disseminated using the \pkg{workflowr} package, usually with nine tabs conveying aspects of the data from collection through several different analyses, and each tab containing approxiately five to thirty tables, plots, or other artifacts including \pkg{trelliscopejs} [@trelliscopejs] environments which may hold hundreds of graphs. By generating many presentation artifacts we are able to address data-drive questions and issues during collaborative sessions and by carefully structuring these elements we allowing all members to participate in the process.

@@

The rest of this paper is structured as follow. The next section goes over basic usage and commentary. This section is meant to convey the basic approach used by the package and shows how to describe an output document using \pkg{listdown}, create a document, and change how the presentation of computational components can be specialized using \pkg{listdown} decorators. With the user accustomed to the package's basic usage, section 3 describes the design of the package. Section 4 goes over advanced usage of the package including adding initialization code to and outputted document as well as how to control chunk-level options. Section 5 provides a simplified case study of how the package is currently being used in clinical trial reporting. Section 6 concludes the paper with a few final remarks on the general types of applications where \pkg{listdown} has been shown effective.

# Basic usage and overview

@@

As a first example, we will consider the a list of visualizations from the anscombe data set. The list is composed of four \pkg{ggplot2} [@wickham2016]elements (named Linear, Non Linear, Outlier Vertical, and Outlier Horizontal) each containing a scatter plot from the famous Anscome Quartet - made available in the \pkg{datasets} package [@R]. From the `computational_components` list, we would like to create a document with four sections with names corresponding to the list names, each containing their respective visualizations.

```{r}
# Use ggplot2 to create the visualizations.
library(ggplot2)

# Load the Anscombe Quartet.
data(anscombe)

# Create the ggplot objects to display.
  computational_components <- list(
    Linear = ggplot(anscombe, aes(x = x1, y = y1)) + geom_point(),
    `Non Linear` = ggplot(anscombe, aes(x = x2, y = y2)) + geom_point(),
    `Outlier Vertical`= ggplot(anscombe, aes(x = x3, y = y3)) + 
      geom_point(),
    `Outlier Horizontal` =  ggplot(anscombe, aes(x = x4, y = y4)) + 
      geom_point())

# Save the file to disk to be read by the output R Markdown document.
saveRDS(computational_components, "comp-comp.rds")
```

## Creating a document with listdown

Creating a document from the `computational_components` requires two steps. First, we will create a `listdown` object that specifies how the `computational_components` object will be loaded into the document,  which libraries and code needs to be included, and how the list elements will be presented in the output R markdown document.

```{r}
library(listdown)

ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"),
               package = "ggplot2")
```

The `ld` object, along with the computational components in the `comp-comp.rds` file are sufficient to to create the sections, subsections, and R chunks of a document. The only other thing requires to create the document is the header. The listdown package currently supports regular R Markdown and \pkg{workflowr} as \code{yml} objects from the \pkg{yaml} package [@yaml]. These objects are stored as named lists in R and are easily modified to accomodate document parameters. A complete document can then be written to the console using the code shown below. It could easily be written to a file for rendering using the `writeLines()` function.

```{r}
doc <- c(
  as.character(ld_rmarkdown_header("Anscombe's Quartet",
                                   author = "Francis Anscombe",
                                   date = "1973")),
  ld_make_chunks(ld))

doc
```

The `listdown()` function provides _document-wide_ R chunk options for displaying computational components. The chunk options are exactly the same as those in the R markdown document and can be used to tailor the default presentation for a variety of needs. The complete set of options can be found in the R Markdown Reference Guide [@rmarkdownref]. As a concrete example, the code used to create present the plots could be hidden in the output document using the following code.

```{r}
ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
               package = "ggplot2",
               echo = FALSE)

#cat(paste(ld_make_chunks(ld)[1:7], collapse = "\n"))
ld_make_chunks(ld)[1:7]
```

## Decorators

The first example is simple in part because the ggplot objects both contains the data we want to display and, at the same time, provides the mechanism for presenting them - rendering them in a graph. However, this is not always the case. The objects being stored in the list of computational components may not translate directly to the presentation in a document. In these cases, a function is needed that takes the list component and returns an object to be displayed. For example, suppose that, along with showing graphs from the Anscombe Quartet, we would like to include the data themselves. We could add the data to the `computational_components` list and then create the document with:

```{r results="as.is"}
computational_components$Data <- anscombe
saveRDS(computational_components, "comp-comp.rds")
ld_make_chunks(ld)[32:36]
```

In this case, the \pkg{listdown} package will show the entire data set as is the default specified. However, suppose we do not want to show the entire data set in the document. This is common, especially when the data set is large and requires too much vertical space in the outputted document resulting in too much or irrelevant data being shown. Instead, we would like to output to an html document where the data is shown in a `datatable` thereby controlling the amount of real-estate needed to present the data and, at the same time, providing the user with interactivity to sort and search the data set.

In \pkg{listdown}, a function or method that implements the presentation of a computational component is referred to as a _decorator_ since if follows the classic decorator pattern described in @gamma1995. A decorator takes the element that will be presented as an argument and returns an object for presentation in the output directory. A decorator is specified using the `decorator` parameter of the `listdown()` function using a named list where the name corresponds to the type and the element correspond to the function or method that will decorate an object of that type. For example, the `anscombe` data set can be decorated with the `DT::datatable()` function [@xie2020] as:

```{r}
ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
               package = c("ggplot2", "DT"),
               decorator = list(data.frame = datatable))

ld_make_chunks(ld)[33:37]
```

List names in the `decorator` argument provide a key to which a function or method is mapped. The underlying decorator resolution is implemented for a given computational component by going through decorator names sequentially to see if the component inherits from the name using the `inherits()` function. The function or method is selected from the corresponding name which the element first inherits from. This means that when customizing the presentation of objects that inherit from a common class, the more abstract classes should appear at the end of the list. This will ensure that specialized classes will be encountered first in the resolution process. It should be noted that an object's type is first checked against the decorator name list and then checked to see if it is a list. This allows a user to both decorate a list and retain `"list"` in its class attributes.

A separate argument, `default_decorator`, allows the user to specify the default decorator for an object whose type does not appear in the `decorator` list. This allows the user to specify any class name for the decorator and avoids a potential type name collision with a default decorator whose name is determined by convention. By default, this argument is set to `identity` but it can be use to not display a computational component by default if the argument is set to `NULL`.

# Design

A \pkg{listdown} object specifies the location of a list of computational components and options for presenting those components in an R Markdown document. The list is a hierarchical data structure that also provides the structure of the outputted document. A corresponding document has two sections "Iris" and "Sepal.Length". The latter has three subsections "Sepal.Width", "Petal.Length", and "Colored". The "Colored" subsection has two sub-subsections, "Sepal.Width" and "Petal.Length".
The structure can be seen using the `ld_cc_dendro()` function.

```{r}
# Create a more hierarchical list of computational components.
comp_comp2 <- list(
  Iris = iris,
  Sepal.Length = list(
    Sepal.Width = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
      geom_point(),
    Petal.Length = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
      geom_point(),
    Colored = list(
      Sepal.Width = ggplot(iris, 
                          aes(x = Sepal.Length, y = Sepal.Width, 
                              color = Species)) + geom_point(),
      Petal.Length = ggplot(iris,
                            aes(x = Sepal.Length, y = Petal.Length, 
                                color = Species)) + geom_point())))

# Create the dendrogram.
ld_cc_dendro(comp_comp2)
```

Both the `ld_cc_dendro()` and `ld_make_chunks()` functions work by recursively descending the computational components list depth-first. If the list containing and element has a name, it is written to the output as a section, subsection, subsubsection, etc. to a return string. If the visited list element is itself a list, then the same procedure is called on the child list through a recursive call. If the element is not a list, then it is outputted inside an R Markdown chunk in the return string using the appropropriate decorator.

# Advanced usage

## Initialization code

The `listdown()` function facilitates the insertion of initialization code through the `init_expr` argument. Code is inserted immediately after the libraries are loaded in the R Markdown document. In general, it is suggested that the number of initial expressions be kept small so that the \proglang{R} Markdown document is easy to read. If a large number of functions are required by the target \proglang{R} Markdown document then they can be put into a file and sourced using the initial expression. As an example, suppose we are creating an html document and presenting data using the `datatable()` function. However, we do not want to include the search capabilities provided. This can be easily accomplished by creating a new function, `datatable_no_search()`, created using the `partial()` function \cite{purrr} to partially apply `list(dom = 't')` to the `options` argument of `datatable`.

```{r}
saveRDS(comp_comp2, "comp-comp2.rds")
ld <- listdown(load_cc_expr = readRDS("comp-comp2.rds"),
               package = c("ggplot2", "DT", "purrr"),
               decorator = list(ggplot = identity,
                                data.frame = datatable_no_search),
               init_expr = {
                 datatable_no_search <- partial(datatable,
                                                options = list(dom = 't'))
                 },
               echo = FALSE)

ld_make_chunks(ld)[2:10]
```

## R code chunk customization

@@

## Controlling options associated with a decorator

The document-wide chunk option specification provides the default chunk options for output documents generated using \pkg{listdown}. However, the presentation of a data object often varies by type. For example, we may want to specify the height and width of a graph, but not a table. This is accomplished in the \pkg{listdown} package when a `listdown` object is created using the `decorator_chunk_opts` option in the `listdown()` function. For example, associating all `ggplot` objects with \proglang{R} chunks having a width of 100 and a height of 200 can be accomplished with the following code and it can be seen that only chunk options associated with a plot are modified.

```{r}
ld <- listdown(load_cc_expr = readRDS("comp-comp2.rds"),
               package = c("ggplot2", "DT", "purrr"),
               decorator_chunk_opts = 
                 list(ggplot = list(fig.width = 100,
                                    fig.height = 200)),
               init_expr = {
                 datatable_no_search <- partial(datatable,
                                                options = list(dom = 't'))
                 },
               echo = FALSE)

ld_make_chunks(ld)[c(12:16, 19:24)]
```

## Controlling chunk-level options

>>>>

\documentclass[
]{jss}

\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Michael Kane\\Yale University
}
\title{On the Programmatic Generation of Reproducible Documents}

\Plainauthor{Michael Kane}
\Plaintitle{On the Programmatic Generation of Reproducible Documents}
\Shorttitle{Programmatic Generation of Reproducible Documents}

\Abstract{
The abstract of the article.
}

\Keywords{reproducibility, reproducible documents, data presentation}
\Plainkeywords{reproducibility, reproducible documents, data
presentation}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Michael Kane\\
  Yale University\\
  60 College StreetNew Haven, CT 06510, USA\\
  E-mail: \href{mailto:michael.kane@yale.edu}{\nolinkurl{michael.kane@yale.edu}}\\

  }

% Pandoc header

\usepackage{amsmath}

\begin{document}

\hypertarget{background-and-concepts}{%
\section{Background and concepts}\label{background-and-concepts}}

\proglang{R} Markdown \citep{baumer2014} facilitates the construction of
computationally reproducible documents by allowing authors to insert
\proglang{R} code for data processing, exploration, analysis,
table-making, and visualization directly into structured, electronic
documents. The resulting documents are made up of these chunks of
\proglang{R} code, which we will refer to as \emph{computational
components} since they are generated by computational means, as well as
\emph{narrative components}, which (in scientific writing) is prose
intended to contextualize computational components, provide background,
define goals, establish themes, and convey the results of the document.
The integration of narrative and computational components was identified
as ``Literate Programming'' by \citet{knuth1984} and software tools,
like Sweave \citep{leisch2002}, have supported this functionality for
almost two decades.

\proglang{R} Markdown is particularly popular with its success likely
being driven by two factors. The first is the relative ease with which
these documents can be constructed. While \LaTeX ~is more expressive, it
is relatively technical and requires an investment in time to become
proficient. By contrast \proglang{R} Markdown documents are easier to
create and format and, when the document is used to create \LaTeX,
formatting can be passed through to the underlying .tex file. The second
factor driving adoption is likely its support for creating
\emph{modifiable} documents, namely Microsoft Word documents.
Researchers and analysts, especially those creating applied statistical
analyses, often collaborate with domain experts with less technical
knowledge. In these cases, the analyst focuses on creating the
computational components and narrative components related to results and
interpretation. After this initial document is created, the domain
expert is free to develop narrative components directly in the document
without needing to go through the analyst.

Since computational components are, by definition, computationally
derived objects and R Markdown is a well-defined standard, it is
possible to programmatically create \proglang{R} Markdown documents with
computational components, which is the focus of this paper. However,
before proceeding down this avenue, we would like to highlight two
fundamental limitations to automated R Markdown document construction.
First and foremost, without narrative components a document has very
little context. Quantitative analyses require research questions,
hypotheses, reviews, interpretations, and conclusions. Computational
components are necessary but generally not sufficient for constructing
an analysis. Second, it is difficult if not impossible to construct
computational components for an arbitary analyses. Analyses themselves
have context and are built with a set of assumptions and goals.

This is not to say the programmatic creation of \proglang{R} Markdown
documents is unwarranted or not useful. In fact, it has at least two
appealing characteristics. First, it allows us to distinguish the
presentation of analytical results from other steps in a data science or
data processing pipeline. These other steps including cleaning and
analysis often require their own environment and configuration with
requirements very different than the computational needs of creating a
presentation. By separating these components each can be developed
independently. At the same time, by specifying a contract for the output
of those objects, we can establish a consistent means by which processed
data can be passed to systems for presentating those data in a
structured way. The second reason for programmatic creation of R
Markdown documents is convenience. In collaborative environments,
especially in the early stages, large numbers of graphs and tables are
generated and discussed. By collecting these artifacts and structuring
them consistently, we can quickly iterate upon and restructure the
resulting documents to more clearly present the data without needing to
spend time on the creation of the presentation document.

The \pkg{listdown} package provides functions to programmatically create
\proglang{R} Markdown files from named lists. It is intended for data
analysis pipelines where the presentation of the results is separated
from their creation. For this use case, a data processing (or analysis)
is performed and the results are provided in a single named list,
organized heirarchically. List element names denote sections,
subsections, subsubsection, etc. and the list elements contain the data
structure to be presented including graphs and tables. The package has
native support for \pkg{workflowr} \citep{blischak2019}, pdf, word, or
html document along with functions allowing a user to easily extend to
other types of supported documents. The goal of the package is to create
a documents with all tables and visualization that will appear
(computational components). This serves as a starting point from which a
user can organize outputs, describe a study, discuss results, and
provide conclusions (narrative components).

@@

In this case, data presentation can be fixed across trials. This is
especially true in the initial stages, which focus on patient
demographics and enrollment. This approach has made it convenient for
our group to quickly generate standardize and complete reports for
multiple trials concurrently. To date, we have used listdown to report
on five clinical trials, with another two currently in process. Results
are disseminated using the \pkg{workflowr} package, usually with nine
tabs conveying aspects of the data from collection through several
different analyses, and each tab containing approxiately five to thirty
tables, plots, or other artifacts including \pkg{trelliscopejs}
\citep{trelliscopejs} environments which may hold hundreds of graphs. By
generating many presentation artifacts we are able to address data-drive
questions and issues during collaborative sessions and by carefully
structuring these elements we allowing all members to participate in the
process.

@@

The rest of this paper is structured as follow. The next section goes
over basic usage and commentary. This section is meant to convey the
basic approach used by the package and shows how to describe an output
document using \pkg{listdown}, create a document, and change how the
presentation of computational components can be specialized using
\pkg{listdown} decorators. With the user accustomed to the package's
basic usage, section 3 describes the design of the package. Section 4
goes over advanced usage of the package including adding initialization
code to and outputted document as well as how to control chunk-level
options. Section 5 concludes the paper with a few final remarks.

\hypertarget{basic-usage-and-overview}{%
\section{Basic usage and overview}\label{basic-usage-and-overview}}

@@

As a first example, we will consider the a list of visualizations from
the anscombe data set. The list is composed of four \pkg{ggplot2}
\citep{wickham2016}elements (named Linear, Non Linear, Outlier Vertical,
and Outlier Horizontal) each containing a scatter plot from the famous
Anscome Quartet - made available in the \pkg{datasets} package
\citep{R}. From the \texttt{computational\_components} list, we would
like to create a document with four sections with names corresponding to
the list names, each containing their respective visualizations.

\begin{CodeChunk}

\begin{CodeInput}
R> # Use ggplot2 to create the visualizations.
R> library(ggplot2)
R> 
R> # Load the Anscombe Quartet.
R> data(anscombe)
R> 
R> # Create the ggplot objects to display.
R>   computational_components <- list(
R+     Linear = ggplot(anscombe, aes(x = x1, y = y1)) + geom_point(),
R+     `Non Linear` = ggplot(anscombe, aes(x = x2, y = y2)) + geom_point(),
R+     `Outlier Vertical`= ggplot(anscombe, aes(x = x3, y = y3)) + 
R+       geom_point(),
R+     `Outlier Horizontal` =  ggplot(anscombe, aes(x = x4, y = y4)) + 
R+       geom_point())
R> 
R> # Save the file to disk to be read by the output R Markdown document.
R> saveRDS(computational_components, "comp-comp.rds")
\end{CodeInput}
\end{CodeChunk}

\hypertarget{creating-a-document-with-listdown}{%
\subsection{Creating a document with
listdown}\label{creating-a-document-with-listdown}}

Creating a document from the \texttt{computational\_components} requires
two steps. First, we will create a \texttt{listdown} object that
specifies how the \texttt{computational\_components} object will be
loaded into the document, which libraries and code needs to be included,
and how the list elements will be presented in the output R markdown
document.

\begin{CodeChunk}

\begin{CodeInput}
R> library(listdown)
R> 
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"),
R+                package = "ggplot2")
\end{CodeInput}
\end{CodeChunk}

The \texttt{ld} object, along with the computational components in the
\texttt{comp-comp.rds} file are sufficient to to create the sections,
subsections, and R chunks of a document. The only other thing requires
to create the document is the header. The listdown package currently
supports regular R Markdown and \pkg{workflowr} as \code{yml} objects
from the \pkg{yaml} package \citep{yaml}. These objects are stored as
named lists in R and are easily modified to accomodate document
parameters. A complete document can then be written to the console using
the code shown below. It could easily be written to a file for rendering
using the \texttt{writeLines()} function.

\begin{CodeChunk}

\begin{CodeInput}
R> doc <- c(
R+   unlist(strsplit(as.character(ld_rmarkdown_header("Anscombe's Quartet",
R+                                author = "Francis Anscombe",
R+                                date = "1973")),
R+                   "\n", fixed = TRUE)),
R+   ld_make_chunks(ld))
R> 
R> doc
\end{CodeInput}

\begin{CodeOutput}
 [1] "---"                                  
 [2] "title: Anscombe's Quartet"            
 [3] "author: Francis Anscombe"             
 [4] "date: '1973'"                         
 [5] "output: html_document"                
 [6] "---"                                  
 [7] ""                                     
 [8] "```{r}"                               
 [9] "library(ggplot2)"                     
[10] ""                                     
[11] "cc_list <- readRDS(\"comp-comp.rds\")"
[12] "```"                                  
[13] ""                                     
[14] "# Linear"                             
[15] ""                                     
[16] "```{r}"                               
[17] "cc_list$Linear"                       
[18] "```"                                  
[19] ""                                     
[20] "# Non Linear"                         
[21] ""                                     
[22] "```{r}"                               
[23] "cc_list$`Non Linear`"                 
[24] "```"                                  
[25] ""                                     
[26] "# Outlier Vertical"                   
[27] ""                                     
[28] "```{r}"                               
[29] "cc_list$`Outlier Vertical`"           
[30] "```"                                  
[31] ""                                     
[32] "# Outlier Horizontal"                 
[33] ""                                     
[34] "```{r}"                               
[35] "cc_list$`Outlier Horizontal`"         
[36] "```"                                  
\end{CodeOutput}
\end{CodeChunk}

The \texttt{listdown()} function provides \emph{document-wide} R chunk
options for displaying computational components. The chunk options are
exactly the same as those in the R markdown document and can be used to
tailor the default presentation for a variety of needs. The complete set
of options can be found in the R Markdown Reference Guide
\citep{rmarkdownref}. As a concrete example, the code used to create
present the plots could be hidden in the output document using the
following code.

\begin{CodeChunk}

\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
R+                package = "ggplot2",
R+                echo = FALSE)
R> 
R> #cat(paste(ld_make_chunks(ld)[1:7], collapse = "\n"))
R> ld_make_chunks(ld)[1:7]
\end{CodeInput}

\begin{CodeOutput}
[1] ""                                     
[2] "```{r echo = FALSE}"                  
[3] "library(ggplot2)"                     
[4] ""                                     
[5] "cc_list <- readRDS(\"comp-comp.rds\")"
[6] "```"                                  
[7] ""                                     
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{decorators}{%
\subsection{Decorators}\label{decorators}}

The first example is simple in part because the ggplot objects both
contains the data we want to display and, at the same time, provides the
mechanism for presenting them - rendering them in a graph. However, this
is not always the case. The objects being stored in the list of
computational components may not translate directly to the presentation
in a document. In these cases, a function is needed that takes the list
component and returns an object to be displayed. For example, suppose
that, along with showing graphs from the Anscombe Quartet, we would like
to include the data themselves. We could add the data to the
\texttt{computational\_components} list and then create the document
with:

\begin{CodeChunk}

\begin{CodeInput}
R> computational_components$Data <- anscombe
R> saveRDS(computational_components, "comp-comp.rds")
R> ld_make_chunks(ld)[32:36]
\end{CodeInput}

\begin{CodeOutput}
[1] "# Data"              
[2] ""                    
[3] "```{r echo = FALSE}"
[4] "cc_list$Data"        
[5] "```"                
\end{CodeOutput}
\end{CodeChunk}

In this case, the \pkg{listdown} package will show the entire data set
as is the default specified. However, suppose we do not want to show the
entire data set in the document. This is common, especially when the
data set is large and requires too much vertical space in the outputted
document resulting in too much or irrelevant data being shown. Instead,
we would like to output to an html document where the data is shown in a
\texttt{datatable} thereby controlling the amount of real-estate needed
to present the data and, at the same time, providing the user with
interactivity to sort and search the data set.

In \pkg{listdown}, a function or method that implements the presentation
of a computational component is referred to as a \emph{decorator} since
if follows the classic decorator pattern described in \citet{gamma1995}.
A decorator takes the element that will be presented as an argument and
returns an object for presentation in the output directory. A decorator
is specified using the \texttt{decorator} parameter of the
\texttt{listdown()} function using a named list where the name
corresponds to the type and the element correspond to the function or
method that will decorate an object of that type. For example, the
\texttt{anscombe} data set can be decorated with the
\texttt{DT::datatable()} function \citep{xie2020} as:

\begin{CodeChunk}

\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
R+                package = c("ggplot2", "DT"),
R+                decorator = list(data.frame = datatable))
R> 
R> ld_make_chunks(ld)[33:37]
\end{CodeInput}

\begin{CodeOutput}
[1] "# Data"                  
[2] ""                       
[3] "```{r}"                  
[4] "datatable(cc_list$Data)"
[5] "```"                    
\end{CodeOutput}
\end{CodeChunk}

List names in the \texttt{decorator} argument provide a key to which a
function or method is mapped. The underlying decorator resolution is
implemented for a given computational component by going through
decorator names sequentially to see if the component inherits from the
name using the \texttt{inherits()} function. The function or method is
selected from the corresponding name which the element first inherits
from. This means that when customizing the presentation of objects that
inherit from a common class, the more abstract classes should appear at
the end of the list. This will ensure that specialized classes will be
encountered first in the resolution process. It should be noted that an
object's type is first checked against the decorator name list and then
checked to see if it is a list. This allows a user to both decorate a
list and retain \texttt{"list"} in its class attributes.

A separate argument, \texttt{default\_decorator}, allows the user to
specify the default decorator for an object whose type does not appear
in the \texttt{decorator} list. This allows the user to specify any
class name for the decorator and avoids a potential type name collision
with a default decorator whose name is determined by convention. By
default, this argument is set to \texttt{identity} but it can be use to
not display a computational component by default if the argument is set
to \texttt{NULL}.

\hypertarget{design}{%
\section{Design}\label{design}}

A \pkg{listdown} object specifies the location of a list of
computational components and options for presenting those components in
an R Markdown document. The list is a hierarchical data structure that
also provides the structure of the outputted document. A corresponding
document has two sections ``Iris'' and ``Sepal.Length''. The latter has
three subsections ``Sepal.Width'', ``Petal.Length'', and ``Colored''.
The ``Colored'' subsection has two sub-subsections, ``Sepal.Width'' and
``Petal.Length''. The structure can be seen using the
\texttt{ld\_cc\_dendro()} function.

\begin{CodeChunk}

\begin{CodeInput}
R> # Create a more hierarchical list of computational components.
R> comp_comp2 <- list(
R+   Iris = iris,
R+   Sepal.Length = list(
R+     Sepal.Width = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
R+       geom_point(),
R+     Petal.Length = ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
R+       geom_point(),
R+     Colored = list(
R+       Sepal.Width = ggplot(iris, 
R+                           aes(x = Sepal.Length, y = Sepal.Width, 
R+                               color = Species)) + geom_point(),
R+       Petal.Length = ggplot(iris,
R+                             aes(x = Sepal.Length, y = Petal.Length, 
R+                                 color = Species)) + geom_point())))
R> 
R> # Create the dendrogram.
R> ld_cc_dendro(comp_comp2)
\end{CodeInput}

\begin{CodeOutput}

comp_comp2
  |-- Iris
  |  o-- object of type(s):data.frame
  o-- Sepal.Length
   |-- Sepal.Width
   |  o-- object of type(s):gg ggplot
   |-- Petal.Length
   |  o-- object of type(s):gg ggplot
   o-- Colored
    |-- Sepal.Width
    |  o-- object of type(s):gg ggplot
    o-- Petal.Length
       o-- object of type(s):gg ggplot
\end{CodeOutput}
\end{CodeChunk}

Both the \texttt{ld\_cc\_dendro()} and \texttt{ld\_make\_chunks()}
functions work by recursively descending the computational components
list depth-first. If the list containing and element has a name, it is
written to the output as a section, subsection, subsubsection, etc. to a
return string. If the visited list element is itself a list, then the
same procedure is called on the child list through a recursive call. If
the element is not a list, then it is outputted inside an R Markdown
chunk in the return string using the appropropriate decorator.

\hypertarget{initialization-code}{%
\section{Initialization code}\label{initialization-code}}

The \texttt{listdown()} function facilitates the insertion of
initialization code through the \texttt{init\_expr} argument. Code is
inserted immediately after the libraries are loaded in the R Markdown
document. In general, it is suggested that the number of initial
expressions be kept small so that the \proglang{R} Markdown document is
easy to read. If a large number of functions are required by the target
\proglang{R} Markdown document then they can be put into a file and
sourced using the initial expression. As an example, suppose we are
creating an html document and presenting data using the
\texttt{datatable()} function. However, we do not want to include the
search capabilities provided. This can be easily accomplished by
creating a new function, \texttt{datatable\_no\_search()}, created using
the \texttt{partial()} function \cite{purrr} to partially apply
\texttt{list(dom\ =\ \textquotesingle{}t\textquotesingle{})} to the
\texttt{options} argument of \texttt{datatable}.

\begin{CodeChunk}

\begin{CodeInput}
R> saveRDS(comp_comp2, "comp-comp2.rds")
R> ld <- listdown(load_cc_expr = readRDS("comp-comp2.rds"),
R+                package = c("ggplot2", "DT", "purrr"),
R+                decorator = list(ggplot = identity,
R+                                 data.frame = datatable_no_search),
R+                init_expr = {
R+                  datatable_no_search <- partial(datatable,
R+                                                 options = list(dom = 't'))
R+                  },
R+                echo = FALSE)
R> 
R> ld_make_chunks(ld)[2:10]
\end{CodeInput}

\begin{CodeOutput}
[1] "```{r echo = FALSE}"                                                   
[2] "library(ggplot2)"                                                      
[3] "library(DT)"                                                           
[4] "library(purrr)"                                                        
[5] ""                                                                      
[6] "cc_list <- readRDS(\"comp-comp2.rds\")"                                
[7] ""                                                                      
[8] "datatable_no_search <- partial(datatable, options = list(dom = \"t\"))"
[9] "```"                                                                   
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{r-code-chunk-customization}{%
\section{R code chunk customization}\label{r-code-chunk-customization}}

@@

\hypertarget{controlling-options-associated-with-a-decorator}{%
\subsection{Controlling options associated with a
decorator}\label{controlling-options-associated-with-a-decorator}}

The document-wide chunk option specification provides the default chunk
options for output documents generated using \pkg{listdown}. However,
the presentation of a data object often varies by type. For example, we
may want to specify the height and width of a graph, but not a table.
This is accomplished in the \pkg{listdown} package when a
\texttt{listdown} object is created using the
\texttt{decorator\_chunk\_opts} option in the \texttt{listdown()}
function. For example, associating all \texttt{ggplot} objects with
\proglang{R} chunks having a width of 100 and a height of 200 can be
accomplished with the following code and it can be seen that only chunk
options associated with a plot are modified.

\begin{CodeChunk}

\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp2.rds"),
R+                package = c("ggplot2", "DT", "purrr"),
R+                decorator_chunk_opts = 
R+                  list(ggplot = list(fig.width = 100,
R+                                     fig.height = 200)),
R+                init_expr = {
R+                  datatable_no_search <- partial(datatable,
R+                                                 options = list(dom = 't'))
R+                  },
R+                echo = FALSE)
R> 
R> ld_make_chunks(ld)[c(12:16, 19:24)]
\end{CodeInput}

\begin{CodeOutput}
 [1] "# Iris"                                                
 [2] ""                                                      
 [3] "```{r echo = FALSE}"                                   
 [4] "cc_list$Iris"                                          
 [5] "```"                                                   
 [6] ""                                                      
 [7] "## Sepal.Width"                                        
 [8] ""                                                      
 [9] "```{r echo = FALSE, fig.width = 100, fig.height = 200}"
[10] "cc_list$Sepal.Length$Sepal.Width"                      
[11] "```"                                                   
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{controlling-chunk-level-options}{%
\subsection{Controlling chunk-level
options}\label{controlling-chunk-level-options}}
